# Mortality

```{r mortality graphs}

# Create dead dataset

# Dead species totals
deadList<-
  X %>% 
  filter(Status=="Dead") %>% 
  mutate(ID=paste(Plot,TreeID,sep="-")) %>% 
  group_by(ID) %>% 
  summarise_all(first)


dead<-
  X %>% 
  filter(Status=="Dead") %>% 
  mutate(ID=paste(Plot,TreeID,sep="-")) %>% 
  mutate(Species=fct_recode(Species,`Douglas-fir`="Fd")) %>% 
  group_by(ID) %>% 
  summarise_all(first) %>% 
  ungroup() %>% 
  mutate(BA=d2ba(DBH)) %>% # convert to basal area per hectare 
  group_by(BA.Target,DBH.class,Year) %>% 
  summarise(BA=sum(BA),
            k=n()) 

dead.Summ<-
  dead %>% 
  ungroup() %>% 
  group_by(BA.Target) %>% 
  summarise(BA=sum(BA)/sum(dead$BA)*100,
            k=sum(k)/sum(dead$k)*100) %>% 
  mutate_if(is.numeric,round,1)

p1<-
  dead %>% 
  ggplot(aes(x=DBH.class,y=k,fill=Year))+
  geom_bar(stat="identity")+
  ylab("Total tree mortality (stems)")+
  facet_wrap(~BA.Target)+
  xlab("")+
  theme(axis.text.x = element_blank())

p2<-
  dead %>% 
  ggplot(aes(x=DBH.class,y=BA,fill=Year))+
  geom_bar(stat="identity")+
  ylab("Total basal area mortality (m²)")+
  facet_wrap(~BA.Target)

mort.plot<-cowplot::plot_grid(p1, p2, ncol = 1, align = "v")
figNums("mort.plot",display=FALSE)


```

A total of `r deadList %>% nrow()` trees (`r deadList %>% filter(Species=="Fir") %>% nrow()` fir, `r deadList %>% filter(Species=="Spruce") %>% nrow()` spruce, and `r deadList %>% filter(Species=="Fd") %>% nrow()` Douglas-fir) died from 1992 to 2019.  No tree mortality was noted during the 1992 and 1994 data collection.  By treatment unit, the control unit had the highest mortality, in terms of trees and basal area (
`r dead.Summ %>% filter(BA.Target=="TU:Control") %>% pull("BA")` and 
`r dead.Summ %>% filter(BA.Target=="TU:Control") %>% pull("BA")`% of tree mortality, respectively).  The high-removal treatment unit had the lowest mortality by number of trees and basal area (
`r dead.Summ %>% filter(BA.Target=="TU:10") %>% pull("BA")` and 
`r dead.Summ %>% filter(BA.Target=="TU:10") %>% pull("BA")`% of tree mortality, respectively).  Tree mortality was concentrated in the smaller diameter classes in the control and low-removal treatment unit, and most mortality occured during the most recent measurement period of 2009-2019 (
`r figNums("mort.plot",display="cite")`).


```{r display mortality plot}
mort.plot
```

`r figNums("mort.plot",caption="Total tree mortality recorded in plots from 1992-2019, presented as total stems (top panel) and basal area (bottom panel).")`

---------------------------

From Mike's email:
Eventually as BA stand density increases, density dependent mortality kicks in, because there is only so much growing space or access to site resources to go around. This can occur with the stand (conceptually or hypothetically)  in two possible ways: (a) “classic” suppression of smaller tree classes through asymmetric competition of bigger trees shading out smaller ones, or (b) high stand densities suppressing the vigor of the whole stand through various means, suggesting more symmetric competition – i.e. the idea that small trees compete for growing space with big ones as well. Asymmetric competition is generally accepted by most foresters and silviculturists; symmetric competition less so, and is often harder to demonstrate. 

*For the discussion: it seems that we're seeing mortality increasing in the plots, especially the control unit over the last 11 years.  Does this reflect the density-dependent mortality that Mike refers to above?.*

More from Mike on mortality: 

The thing about unevenaged or complex stand management, esp in ‘wilder’ BC stands, is that we are working in stands with a long legacy of pre-existing ecosystem processes before we enter the stand to manipulate them. Clearcutting liquidates these legacies for the most part, and traditional silviculture and GY has not had to deal with mortality much, other than clearly density dependent suppression of smaller dbh classes. [Side issue: most BC experience and data with unevenaged stands has been in IDF, which have been fire excluded on dry sites, and unevenaged structures allowed to develop. We must be careful about comparing Summit data to IDF precedents, because of substantial differences in ecology.]

And we are working a more extensive level of management in the SBS and ESSF,  where our stands are older, and not manicured nor vacuumed up for mortality and blowdown. From an ecosystem management point of view, that Is probably a good thing. 

So warts and all, I think mortality is an inherent characteristic of the types of spruce-balsam stands and management we are studying. Expecting and documenting mortality was built into the EP 1162 working plan. The treatment effects we anticipate are the net stand growth = gross growth minus mortality. 

Conceptually, tree mortality following the 1991/92 harvest (and subsequent sanitation felling)  functions as a BA removal from the live-tree pool. In most plots, BA of trees that die within a measurement period is subsumed or overridden by the total BA growth of the rest of the trees during that period. 

Where the periodic data trends get messier is where single big trees or a cluster of trees gets killed by something, resulting in a pulse of BA mortality. Examples: The great big Fdi in Plot 18, or notably, about 3-4 big Sx in Plot 3 that got nailed by spruce beetle. 

In our results and discussion:
1.	We can report on what stand growth and development trends seem to be clearly and unambiguously  influenced by initial BA conditions, and the magnitude of these treatment effects. 
2.	A component of tree mortality (e.g. certain size classes) may be density dependent. 
3.	A component of tree mortality may be density-independent (a.k.a stochastic or chaotic). 
4.	By reporting on all the above, we are providing managers with results that integrate silvicultural expectations observations from real-life “messy” stands which will experience all of the above. By keeping our eye on overall stand growth and performance following treatment, we can inform / reassure managers that individual tree mortality events are not the end of the world, but rather need to be viewed in the context of overall stand dynamics and growth expectations. 