---
title: "EP1162_Prelim_Data_Screening"
author: "Hardy Griesbauer"
date: "18/02/2020"
output:
  html_document: default
  word_document: default
---

```{r setup, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(message=FALSE, warning = FALSE)
library(tidyverse)
library(readxl)
library(here)

##
load(here("data","ep1162_Data.RData"))

```

## Preliminary screening
I did some preliminary screening, to confirm data entry and detect any potential outliers.  Some findings below.


## Outliers
Quick boxplot to confirm DBH and height measurements 

#### DBH

```{r echo=FALSE}

dat %>% 
  ggplot()+
  aes(group=Species,y=DBH,fill=Species) +
  geom_boxplot()


```

DBH measurements look reasonable, and no major outliers.  Note that there is a tree with species recorded as "l".


#### Height

```{r echo=FALSE}

dat %>% 
  ggplot()+
  aes(group=Species,y=Height,fill=Species) +
  geom_boxplot()


```
Again, measurements look reasonable here.

### Data entry

One tree has species recorded as "l".  I assume that it's SAF, but might be worthwhile to verify with the field cards?


```{r echo=FALSE}

dat %>% 
  filter(Species=="l") %>% 
  knitr::kable()

```

*ALRF staff confirmed that the species is "Bl"*, so we'll change the species code accordingly in the next line of code:

```{r}
dat<-
  dat %>% 
  mutate(Species=replace(Species,Species=="l","Bl"))

```


There are also two trees recorded with id=631:

```{r echo=FALSE}
# Look at entries where tree number is duplicated within plot
dat %>% 
  add_count(Plot,TreeID,Year) %>% # get a number by this grouping
  filter(n>1) %>% 
  drop_na(TreeID) %>% 
  filter(TreeID!="New") %>% 
    knitr::kable()
  
```

Looks like it is the same tree (exact same DBH and Height measured in 2019), so assume it's safe to delete the duplicated entry?

*Confirmed that this tree was duplicated by mistake.  We will remove one entry.*

```{r}
dat<-
  dat %>% 
  filter(!(Plot==10 & TreeID==631 & Sector=="NA")) # remove entry

```


There are also quite a few trees with their ID recorded as "New".  Not an issue from an analysis point of view,

### Decreasing DBH 
I also screened trees that had a smaller DBH measurements than a previous measurement at some point.  Below are some examples:

```{r echo=FALSE}
# Look at entries where recorded DBH is less from one measurement to the other

dbhDecrease<-
  dat %>% 
  filter(TreeID!="New") %>% 
  filter(TreeID!=631) %>% # this is a duplicated tree
  pivot_wider(id_cols=c(Plot,TreeID),names_from="Year",values_from="DBH") %>% 
  mutate(diff.2019=`2019`-`2009`) %>% 
  mutate(diff.2009=`2009`-`1997`) %>% 
  mutate(diff.1997=`1997`-`1994`) %>% 
  mutate(diff.1994=`1994`-`1992`) %>% 
  filter(diff.2019<0 | diff.2009<0 | diff.1997<0 | diff.1994<0) %>% 
  dplyr::select(Plot,TreeID) %>% # select these trees with decreasing DBH measurements
  left_join(dat,by=c("Plot","TreeID")) %>% 
  mutate_at(vars(contains("DBH")),round,1) %>% # round to one decimal place
  dplyr::select(Plot:DBH)

  
  # Show example rows
  dbhDecrease %>% 
  slice(1:20) %>% 
  knitr::kable()
  
  


```

In most cases, the differences are very small, and likely can be atrributed to slight measurement differences between years.  

In total, there are `r dbhDecrease %>% count(TreeID) %>% nrow()` trees with this situation.  

*****
## Trees without monotonically increasing DBH measurements
After discussion, we will handle trees without monotonically increasing DBH measurements as follows:

- Where DBH decreases occur in dead trees or during periods where trees died, we will treat these decreases as likely due to wood shrinkage/rot associated with mortality.
- Where DBH decreases occur during periods where trees were live, we will treat these decreases as a measurement error.  We will use linear interpolation to correct these errors.

Code below.

### Trees with decreasing DBH over dead periods

```{r}
# Step one: identify periods where DBH decrease occured during a period when tree waslive

dbhDecrease %>% 
  filter(Status=="Dead") %>% # filter trees with decreasing DBH that had dead status at any point
  left_join(dat,by=c("Plot","TreeID")) %>%  # join with dataset
  dplyr::select(Plot,TreeID,Year=Year.y,Status=Status.y,DBH=DBH.y)



```

A visual inspection reveals that the following trees had DBH decreases during periods where the tree was dead:

- Plot 3, Tree 457
- Plot 10, Tree 409

These trees should be filtered from further adjustment.

```{r}
# Filter out trees with DBH decrease with dead periods
dbhDecrease<-
  dbhDecrease %>% 
  filter(!(Plot==3 & TreeID==457)) %>% 
  filter(!(Plot==10 & TreeID==409))

```

### Trees with decreasing DBH over live periods
We will make adjustments on a tree by tree basis.

First, reshape the dbhDecrease dataframe to make it easier to see DBH by year, then save it to a .csv file for manual editing.

*I set the following chunk not to run, because we only want to do this once*

```{r eval=FALSE}
dbhEdit1<-
  dbhDecrease%>% 
  dplyr::select(-Status) %>% # Don't need this column anymore
  pivot_wider(names_from="Year",values_from="DBH") %>% 
  write.csv(file="dbh_for_edit.csv",row.names=FALSE)

```

One thing to note: the following trees had fallen as of 2019, which means we should not interpolate their 2019 values:


```{r}
dbhDecrease %>% 
  filter(Year==2019 & is.na(DBH))
```

*The following chunk won't run, because we only want to do this once*
```{r eval=FALSE}
dbhEdit2<-
  readr::read_csv(here::here("data-raw","dbh_for_edit.csv"))

# save this file so we can go back to it
save(dbhEdit2,file=paste("editedDBH",Sys.Date(),"RData",sep="."))

```

Also, there are trees that were <7.5cm DBH during the 1992 or 1994 measurements.  DBH values for those trees/periods should not be interpolated either.

For values that shouldn't be interpolated, I entered 'DoNotReplace' in the corresponding cell.

Values to be replaced by interpolation had 'Interpolate' entered into their cell.



```{r}
# Load tree data with values to replace
load(here::here("data-raw","editedDBH.2020-02-27.RData"))

# Show data with replacements
dbhEdit2 

```

Now we can interpolate the missing values

```{r}
library(imputeTS)
library(padr)
library(lubridate)
library(Hmisc)

# Create a vector of years that are not to be interpolated
DoNotReplace<-  
  dbhEdit2 %>% 
  pivot_longer(cols=`1992`:`2019`,names_to="Year",values_to="DBH") %>% 
  mutate(TreeID=factor(TreeID)) %>% 
  mutate(Year=as.numeric(Year)) %>% 
  filter(DBH=="DoNotReplace")

# Interpolation
dbhEdit3<-
  dbhEdit2 %>% 
  pivot_longer(cols=`1992`:`2019`,names_to="Year",values_to="DBH") %>% 
  mutate(DBH=as.numeric(DBH)) %>% 
  mutate(Year=paste("01-01",Year,sep="-")) %>% 
  mutate(Year=dmy(Year)) %>% 
  mutate(TreeID=factor(TreeID)) %>% 
  group_by(TreeID) %>% 
  pad() %>% # pad out time series so we can interpolate
  ungroup() %>% 
  mutate(Year=year(Year)) %>% 
  as.data.frame()

# Define function to predict DBH values
dbhInterp<-function() {
  
  YearList=data.frame(Year=c(1992,1994,1997,2009,2019))  
  
 x<-
   dbhEdit3 %>% 
   lm(DBH~Year,data=.) %>% 
   predict(.,newdata=YearList)
 
 return(data.frame(Year=YearList$Year,DBH.pred=x))
  
}

# Run function by TreeID
dbhPred1<-by(dbhEdit3, dbhEdit3[,"TreeID"],
   function(x) predict(lm(DBH ~ Year, data = x),new=data.frame(Year=c(1992,1994,1997,2009,2019))),simplify=F)

# Bring datasets together 
dbhPred<-data.frame(TreeID=names(dbhPred1),do.call(rbind.data.frame,dbhPred1)) %>% # unpack output from by() function
  setNames(c("TreeID",c(1992,1994,1997,2009,2019))) %>% # rename columns
  pivot_longer(cols=`1992`:`2019`,names_to="Year",values_to = "DBH.Pred") %>% # pivot longer 
  mutate(Year=as.numeric(Year)) %>% 
  
  # Join with dbhEdit3
  left_join(dbhEdit3,by=c("TreeID","Year")) %>% 
  
  # Join with doNot Replace
  left_join(DoNotReplace,by=c("TreeID","Year")) %>% 
  
  # Formatting
  dplyr::select(Plot=Plot.x,TreeID,Year,DBH=DBH.x,DBH.Pred,DoNotRep=DBH.y) %>% 
  mutate_at(vars(DBH.Pred),round,1) %>% # round DBH.Pred
  
  # Find and replace
  mutate(DBH.New=DBH) %>% 
  mutate(DBH.New=coalesce(DBH.New,DBH.Pred)) %>% 
  mutate(DBH.New=replace(DBH.New,!is.na(DoNotRep),"NA")) %>% 
  
  # Set new DBH column
  dplyr::select(Plot:Year,DBH=DBH.New) %>% 
  mutate(DBH.Pred=as.numeric(DBH)) %>% 
  dplyr::select(-DBH) %>% 
  mutate(Plot=factor(Plot)) %>% 
  mutate(Year=factor(Year)) %>% 
  mutate(TreeID=factor(TreeID))


```

Now let's test to make sure that all DBH are now monotonic.

```{r}
library(MonoInc)

x<-
  dbhPred %>% 
  drop_na(DBH.Pred) 
  
 mono.x<- by(x$DBH.Pred,x$TreeID,monotonic)

 # One tree still has decreasing DBH
 dbhPred %>% 
   filter(TreeID==as.numeric(names(which(mono.x==FALSE))))
 
```

One tree still has decreasing DBH. 1994 value for this tree is less than 1992.  This is because the linear function predicts values below the 1992 value.  Let's manually replace the 1994 value with interpolation between 1992 and 1997 values.

```{r}

  dbhPred[dbhPred$TreeID==31 & dbhPred$Year==1994,"DBH.Pred"]=22.5

```


*** 
## Now merge DBH values back into the dataset
```{r}
dat<-
  dat %>% 
  left_join(dbhPred,by=c("Plot","TreeID","Year")) %>% 
  arrange(Plot,TreeID,Year) %>% 
  mutate(DBH=coalesce(DBH.Pred,DBH)) %>% 
  mutate_at(vars(DBH),round,1) %>% 
  dplyr::select(-DBH.Pred)


```

# Heights
Now let's look at trees where height measurements aren't monotonically increasing.


```{r}
x<-
  dat %>% 
  drop_na(Height) %>% # drop rows with missing Height values 
  arrange(Plot,TreeID,Year) %>% 
  mutate(id=paste(Plot,TreeID,sep="."))
  
 mono.x<- by(x$Height,x$id,monotonic)

 
  # Trees with decreasing height measurements

  x %>% 
   group_by(id) %>% 
   summarise(n=n()) %>% 
   filter(n>1) %>%  # filter tree records with more than one height measurement
   filter(id%in%names(which(mono.x==FALSE))) %>% 
   left_join(x,by="id") %>% 
   arrange(id) %>% 
   dplyr::select(Plot,TreeID,Species,Year,Height) %>% 
   pivot_wider(names_from="Year",values_from = "Height") 
 

```

42 trees have decreasing DBH.  A visual inspection of the plot cards revealed several trees with broken tops (e.g., Plot 12 Tree 22, and Plot 18 Tree 346).

In almost all other cases, there is too much fluctuation to confidently delete height measurements for interpolation.  I recommend we leave the height measurements as is.

### Remove trees with DBH<7.5cm DBH for that year
We need to remove trees from the record that have less than 7.5cm DBH recorded.

There are `dat %>% filter(DBH<7.5) %>% nrow()` records of trees with less than 7.5cm DBH that will be removed using the following chunk.

```{r}

dat<- # re-assign to dat
  dat %>% 
  filter(DBH>=7.5) # keep trees >= 7.5cm DBH

```

### Fix measurement for single tree
One tree (Plot==11 TreeID==991) had a measurement missed in 1997.  I interpolated using surround values to fix it:

```{r}
# Take a look at tree records
dat %>% filter(Plot==11 & TreeID==991)
  
# add DBH 17.2 (mean of 1994 and 2009 values)
dat<-
  dat %>% 
  add_row(Plot=11,TreeID=991,DBH=17.2,Status="Live",Species="Bl",Sector=1,
    Year=1997,Comments=c("DBH added manually for this year, missing"))

# Confirm that entry worked
dat %>% filter(Plot==11 & TreeID==991)


```



*** 
# FINAL SAVING
We'll save the final dataset and export it as well to .csv for final checks.

```{r}
dat %>% 
  write.csv(file=here::here("data-raw","ep1162_finalData.csv"),
            row.names=F,
            na="")

  save(dat,file=here::here("data","ep1162_Data.RData"))


```


